\begin{abstract}
Tracking long-term capital accumulation and neighborhood-level economic change is frequently constrained by the high-variance sampling noise and transient shocks inherent in survey-based datasets like the American Community Survey (ACS). While aerial computer vision models offer high-frequency spatial tracking, they conventionally suffer from \textit{Structural Cardinality Mismatch}---memorizing the localized topological scale of their training data (e.g., New York's vertical density versus Dallas's horizontal sprawl) and failing to generalize out-of-sample. In this paper, we introduce a Mixed Spatial-Temporal Siamese Network explicitly designed with a dual objective: to map the cross-sectional distribution of structural wealth and to predict the magnitude of longitudinal structural change (gentrification and decay) across a multi-year panel (2008--2024). 

To isolate true physical capital accumulation from transient econometric ``yo-yo'' shocks, we engineer a heavily smoothed, inverse-variance weighted training panel. Rather than parametrically regressing exact income levels, our model acts as a non-parametric ordinal feature extractor trained via a tripartite loss function. By utilizing a Spatial Margin Ranking loss, we intentionally discard topological cardinality, enabling zero-shot out-of-sample generalization to novel cities via direct quantile matching with local survey moments. To prevent the ``vanishing gradient'' problem typical of ordinal rankers tracking temporal change, we introduce a Temporal Dynamic Margin proportional to the severity of true econometric shifts. We validate the model's capacity to internalize continuous magnitudes and localized economic externalities through mid-point (2016) holdout interpolations and adjacent-tract spatial spillover testing. Ultimately, this framework provides a highly robust, universally generalizable tool for mapping both static income distributions and dynamic urban evolution from visual infrastructure alone.
\end{abstract}