\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{geometry}
\geometry{a4paper, margin=1in}

\begin{document}

\section{Methodology: Data Processing and Label Engineering}

To train a computer vision model to detect long-term capital accumulation, we must first isolate true structural neighborhood change from the high-variance sampling noise inherent in American Community Survey (ACS) 5-year estimates. If the network is trained on raw survey data, it risks optimizing for transient economic shocks or survey artifacts rather than physical infrastructure changes. We construct a robust, smoothed longitudinal panel ($t \in \{2009, 2014, 2019, 2024\}$) using the following procedure.

\subsection{Spatial Alignment}
To resolve the Modifiable Areal Unit Problem (MAUP) caused by shifting census boundaries, we define the 2024 census tracts as our fixed spatial geometry. Historical tracts from 2009, 2014 and 2019 are spatially mapped to the 2024 baseline using a Maximum Area Overlap criterion, ensuring a consistent longitudinal spatial unit $i$ across all periods $t$.

\subsection{Log-Normalization and the Delta Method}
Let $Y_{i,t}$ denote the raw Per Capita Income for tract $i$ in year $t$, and let $MOE_{i,t}$ denote the Census-reported 90\% Margin of Error. Because income is right-skewed and we evaluate proportional growth, we apply a natural logarithmic transformation. The standard errors are first extracted from the 90\% confidence interval, $SE(Y_{i,t}) = MOE_{i,t} / 1.645$, and then transformed into logarithmic space utilizing the first-order Taylor series approximation (the Delta Method):
\begin{equation}
SE(\ln Y_{i,t}) \approx \frac{SE(Y_{i,t})}{Y_{i,t}}
\end{equation}

\subsection{Relative Cross-Sectional Ranking}
To strip out city-wide macroeconomic inflation and aggregate growth trends, we convert the logged incomes into cross-sectional Z-scores relative to the city-wide distribution for each specific year. Let $\mu_t$ and $\sigma_t$ be the city-wide mean and standard deviation of $\ln Y_{i,t}$. The relative structural wealth score $Z_{i,t}$ and its corresponding relative standard error $\tilde{SE}_{i,t}$ are defined as:
\begin{equation}
Z_{i,t} = \frac{\ln Y_{i,t} - \mu_t}{\sigma_t}, \quad \tilde{SE}_{i,t} = \frac{SE(\ln Y_{i,t})}{\sigma_t}
\end{equation}

\subsection{Structural Break Identification and Transient Shock Filtering}
To separate structural change from sampling noise, we conduct two-tailed Z-tests on the differences between periods. The test statistic for the difference between year $t$ and year $t-k$ is given by:
\begin{equation}
S_{i, (t, t-k)} = \frac{|Z_{i,t} - Z_{i,t-k}|}{\sqrt{\tilde{SE}_{i,t}^2 + \tilde{SE}_{i,t-k}^2}}
\end{equation}
We establish a binary indicator for statistical significance at the 95\% confidence level ($p < 0.05$) across these temporal pairs (2009–2014, 2014–2019, 2019–2024, 2009–2024). Crucially, we apply a ``Macro Gate'' to filter out transient ``Yo-Yo'' shocks---defined as tracts exhibiting a statistically significant positive jump in one period followed immediately by a significant negative drop in the next, or vice versa. Let $I_{i}^{valid}$ be a boolean indicator equal to 1 if tract $i$ exhibits at least one statistically significant change across any temporal pair and is \textit{not} classified as a transient Yo-Yo shock.

\subsection{Siamese Network Training Labels and Economic Filtering}
A core challenge in assigning longitudinal labels is the phenomenon of spatial economic filtering. As a city grows and new high-wealth developments are constructed, the macro-distribution of relative wealth shifts. Consequently, a stable tract ($I_{i}^{valid} = 0$) that experiences zero physical change may naturally slide downward in its relative ranking (a leftward drift in $Z_{i,t}$) over a multi-year panel simply because the rest of the city outpaced it.

To avoid introducing a mechanical downward bias by artificially masking stable tracts with a static historical average, we explicitly decouple the cross-sectional economic label from the longitudinal physical label. The final econometric label $L_{i,t}$ fed into the Spatial Margin Ranking loss for any tract, regardless of its structural stability, is strictly its true, year-specific relative score:
\begin{equation}
L_{i,t} = Z_{i,t}
\end{equation}
Rather than forcing mathematically identical labels on the dataset, the permanence of the physical capital stock for stable tracts is enforced architecturally. During training, the Siamese network's temporal regularization loss ignores the natural economic label drift of stable tracts and strictly penalizes divergence in their extracted visual embeddings, allowing the model to learn absolute physical permanence while preserving shifting relative economic value.

\end{document}