\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{geometry}
\geometry{a4paper, margin=1in}

\title{Methodological Update: In-Batch Ordinal Pairwise Ranking}
\author{}
\date{}

\begin{document}

\maketitle

\section{Motivation: From Isolated Triplets to In-Batch Ranking}

While the theoretical properties of the decoupled Triplet Margin Ranking Loss ensure economic comparability and prevent longitudinal label drift, constructing isolated $(A, P, N)$ triplets offline introduces severe computational inefficiencies and gradient instability. Passing three full-resolution images through a Vision Transformer to extract a single pairwise gradient discards over 98\% of the viable cross-sectional supervision available within a mini-batch. Furthermore, adaptive margins calibrated on isolated triplets are vulnerable to variance collapse early in training.

To resolve this, we transition the architecture to an \textbf{In-Batch Pairwise Ranking} framework. This approach strictly preserves the theoretical ``Comparability Constraint'' defined in the core methodology—we still only compare buildings within the same year, or the same building across different years—but it maximizes supervision density by computing all valid pairwise constraints simultaneously from a single, unified forward pass.

\section{Batch Construction and Sampling Strategy}

Instead of pre-assembling triplets, the dataloader constructs a stochastically stratified hybrid batch, $\mathcal{B}$, composed of two subsets: a cross-sectional core and a temporal auxiliary set.

\subsection{1. The Cross-Sectional Core ($\mathcal{B}_{cs}$)}
\begin{itemize}
    \item \textbf{Source:} A \texttt{HybridBatchSampler} randomly selects a single anchor year, $t$. It then samples $B_{cs}$ distinct buildings observed in year $t$ from the currently loaded cache shards.
    \item \textbf{Purpose:} To map the cross-sectional ordinal wealth distribution of the city at a fixed point in time.
    \item \textbf{Spatial Exclusion:} During loss computation, pairs drawn from the same census tract are strictly masked out, as American Community Survey (ACS) labels provide no intra-tract variance.
\end{itemize}

\subsection{2. The Temporal Auxiliary Set ($\mathcal{B}_{temp}$)}
\begin{itemize}
    \item \textbf{Source:} For a fraction $\rho_{temp}$ (default 0.20) of the $B_{cs}$ buildings in $\mathcal{B}_{cs}$, the sampler queries the longitudinal panel for observations of the same building in a different year, $t' \neq t$.
    \item \textbf{Stable Twins ($\mathcal{B}_{stable}$):} If building $k$ exists in $t'$ and underwent no structural change ($I_{k}^{valid}(t \to t') = 0$), the image for $t'$ is added to the batch.
    \item \textbf{Changed Twins ($\mathcal{B}_{change}$):} If building $k$ exists in $t'$ but underwent statistically significant structural change ($I_{k}^{valid}(t \to t') = 1$), the image for $t'$ is added to the batch.
    \item \textbf{Purpose:} To anchor the latent continuum over time. Stable twins enforce absolute temporal invariance (L1 penalty), while changed twins enforce directional rank mobility (smooth logistic surrogate).
    \item \textbf{Dynamic Pairing:} Each temporal twin is matched to its cross-sectional anchor at loss-computation time by vectorized DOITT\_ID matching, ensuring that the temporal penalty always references the correct cross-sectional partner score.
\end{itemize}

\section{Mathematical Formulation of the Loss}

A single forward pass encodes the entire batch $\mathcal{B} = \mathcal{B}_{cs} \cup \mathcal{B}_{temp}$ into a vector of scalar scores, $\hat{R}$. The total loss is defined as the decoupled sum of three objectives computed over valid subsets of these scores, plus a variance regularizer.

\subsection{Cross-Sectional Ranking: Smooth Logistic Surrogate}

For the $B_{cs}$ cross-sectional images in year $t$, we enumerate all $\binom{B_{cs}}{2}$ unique, undirected pairs via upper-triangular indices. Let $\mathcal{U}_{valid}$ be the set of unique pairs $(k, l)$ where buildings $k$ and $l$ reside in \textit{different} census tracts.

Let $y_{k,l} = \text{sign}(Z_l - Z_k) \in \{-1, +1\}$ represent the true economic ordering from the ACS. Define the signed agreement $\delta_{k,l} = y_{k,l} \cdot (\hat{R}_l - \hat{R}_k)$, which is positive when the model ranks the pair correctly. The cross-sectional loss is:
\begin{equation}
\mathcal{L}_{cross} = \frac{1}{|\mathcal{U}_{valid}|} \sum_{(k,l) \in \mathcal{U}_{valid}} \text{softplus}\!\left(-\frac{\delta_{k,l}}{T}\right)
\end{equation}
where $\text{softplus}(x) = \log(1 + e^x)$, and $T$ is an adaptive temperature:
\begin{equation}
T = \max\!\left(m_{min},\;\; m_{base} \cdot \sigma_{batch}\right)
\end{equation}
Here $\sigma_{batch} = \text{std}(\hat{R}_{cs})$ is the \textit{detached} standard deviation of the cross-sectional predictions within the current batch. This temperature carries zero economic information—it scales with the model's own output distribution, not ACS label magnitudes—preserving the purely ordinal supervision framework.

For a batch of $B_{cs} = 16$, this yields $16 \times 15 / 2 = 120$ unique ranking constraints per micro-batch, or $120 \times 8 = 960$ constraints per effective optimizer step with gradient accumulation.

\paragraph{Curriculum Learning: Easy-to-Hard Pair Filtering.}
To improve early-training gradient quality, each building is assigned to a score quintile bin $q \in \{1, 2, 3, 4, 5\}$ based on its ACS label. An epoch-dependent curriculum progressively admits harder comparisons:
\begin{itemize}
    \item \textbf{Epochs 0--59:} Only pairs with $|q_k - q_l| \geq 3$ (extreme quintile contrasts).
    \item \textbf{Epochs 60--149:} Pairs with $|q_k - q_l| \geq 2$.
    \item \textbf{Epoch 150+:} Pairs with $|q_k - q_l| \geq 1$ (all inter-quintile pairs).
\end{itemize}
Intra-quintile pairs ($|q_k - q_l| = 0$) are \textit{permanently excluded}: the ACS margin of error makes within-quintile orderings pure noise, and training on them causes variance collapse.

\subsection{Temporal Stability Penalty}
For buildings in the stable temporal subset $\mathcal{B}_{stable}$, the latent human wealth bundle is physically identical between year $t$ and year $t'$. To enforce temporal invariance regardless of shifting city-wide distributions, we apply an unconditional L1 penalty:
\begin{equation}
\mathcal{L}_{stable} = \frac{1}{|\mathcal{B}_{stable}|} \sum_{k \in \mathcal{B}_{stable}} \left| \hat{R}_{k,t} - \hat{R}_{k,t'} \right|
\end{equation}

\subsection{Temporal Change Ranking}
For buildings in the changed temporal subset $\mathcal{B}_{change}$, physical alteration identifies a shift in the human wealth bundle. Let $y_{k, t \to t'} \in \{-1, +1\}$ indicate the direction of the structural change. We supervise this displacement using the same smooth logistic surrogate, with a temperature $T_{change} = \max(m_{min}, \; m_{base} \cdot \sigma')$ where $\sigma'$ is computed over the union of cross-sectional and temporal change scores:
\begin{equation}
\mathcal{L}_{change} = \frac{1}{|\mathcal{B}_{change}|} \sum_{k \in \mathcal{B}_{change}} \text{softplus}\!\left(-\frac{y_{k,t\to t'} \cdot (\hat{R}_{k,t'} - \hat{R}_{k,t})}{T_{change}}\right)
\end{equation}

\subsection{Variance Regularizer}
To prevent the score head from collapsing its output to a near-constant value—a failure mode in which the softplus loss is minimized by shrinking $\sigma_{batch}$ toward zero—we apply a quadratic penalty targeting unit standard deviation on the cross-sectional scores exclusively:
\begin{equation}
\mathcal{L}_{var} = \left(\sigma_{cs} - 1\right)^2
\end{equation}
This is internally consistent: both the temperature $T$ and the variance target reference $\sigma_{cs}$, eliminating any tug-of-war between temporal invariance ($\mathcal{L}_{stable}$, which pushes scores together) and the score spread needed for informative ranking gradients.

\subsection{Total Objective}
The final unified objective aggregates these valid comparisons:
\begin{equation}
\mathcal{L}_{total} = \mathcal{L}_{cross} + \lambda_{s} \cdot \mathcal{L}_{stable} + \lambda_{c} \cdot \mathcal{L}_{change} + \mathcal{L}_{var}
\end{equation}
where $\lambda_s$ and $\lambda_c$ are weighting hyperparameters governing the relative influence of the temporal constraints against the cross-sectional ranking baseline.

\end{document}